\vspace{1.0cm} \LARGE \bf MERS-CoV spillover at the camel-human interface

### 2017-08-02 mers-structure.tex
Text changes, add skygrid as main figure.

Although we find evidence of seasonality in zoonotic spillover timing, no such relationship exists for sizes of human sequence clusters \tbc{reference a figure}.
This is entirely expected, since little seasonality in human behaviour that could facilitate MERS-CoV transmission is expected following an introduction.
Similarly, we do not observe any trend in human sequence cluster sizes over time (Figure \ref{seasonality}B, Spearman $\rho = 0.06$, $p=0.68$), suggesting that MERS-CoV outbreaks in humans are neither growing nor shrinking in size.
This is not surprising either, since MERS-CoV is a camel virus that has to date, experienced little-to-no selective pressure to improve transmissibility between humans.

@@

\subsection*{Recombination shapes MERS-CoV diversity}
Recombination has been shown to occur in all genera of coronaviruses, including MERS-CoV \citep{lai_1985,makino_1986,keck_1988,kottier_1995,herrewegh_1998}. %\tbc{needs citations}.
In order to explore the role of recombination in shaping MERS-CoV genetic diversity we used two recombination detection approaches across partitions of taxa corresponding to inferred MERS-CoV clades.
Both methods rely on sampling parental and recombinant alleles within the alignment, although each quantifies different signals of recombination.
One hallmark of recombination is the ability to carry alleles derived via mutation from one lineage to another, which appear as repeated mutations taking place in the recipient lineage, somewhere else in the tree.
The PHI (pairwise homoplasy index) test quantifies the appearance of these excessive repeat mutations (homoplasies) within an alignment \citep{bruen_simple_2006}.
Another hallmark of recombination is spatial clustering of alleles along the genome, due to how template switching, the primary mechanism of recombination in RNA viruses, occurs.
3Seq relies on the spatial structure of nucleotide similarities between sequence triplets -- two potential parent-donors and one candidate offspring-recipient sequences \citep{boni_exact_2007}.

@@

%%% AR - this paragraph is a bit of a mess. Suggests we can't trust most of the analysis above. If we can't believe the camel tree why do we trust a structured coalescent model?
Conversely, our results strongly suggest that co-infection of humans with distinct lineages of MERS-CoV should be exceedingly rare.
We find little evidence that recombination will interfere with the inference of human outbreak clusters (Figure \ref{recombinant_features}A).
\tbc{Summarize Figure \ref{recombinant_features}A with something like: ``We find that 95\% of human viruses fall within the same introduction event in genomic fragment 1 and in genomic fragment 2.''}
And while we observe evidence of distinct phylogenetic trees from different parts of the MERS-CoV genome (Figure \ref{recombinant_features}B), human clades are minimally affected and large portions of the posterior probability density in both parts of the genome are concentrated in shared clades (Figure \ref{flower}).
Critically, we observe the same source-sink dynamics between camel and human populations in trees constructed from separate genomic fragments as were observed in the original full genome tree (Figures \ref{mcc}, \ref{recombinant_features}B).
Observed departures from strictly clonal evolution suggest that while recombination is an issue for inferring MERS-CoV phylogenies, its effect on the human side of MERS outbreaks is minimal, as expected.
MERS-CoV evolution on the reservoir side, though complicated by recombination, is nonetheless still amenable to phylogenetic methods, in part through limited diversity of the virus in camels (see next section).
In humans MERS-CoV evolution should be far easier to track as the only detectable and problematic recombinants are more likely to arise within the transmission chain, than through human co-infection with distinct MERS-CoV lineages.

@@

@@

For our primary analysis, the MultiTypeTree module \citep{vaughan_efficient_2014} of BEAST v2.4.3 \citep{bouckaert_beast_2014} was used to specify a structured coalescent model with two demes -- humans and camels.
This model estimates 4 parameters: rate of coalescence in human viruses, rate of coalescence in camel viruses, rate of migration from the human deme to the camel deme and rate of migration from the camel deme to the human deme.
Analyses were run on codon position partitioned data with two separate HKY+$\Gamma_{4}$ \citep{hky_1985,yang_1994} nucleotide substitution models specified for codon positions 1+2 and 3.
A relaxed molecular clock with branch rates drawn from a lognormal distribution \citep{drummond_2006} was used to infer the evolutionary rate from date calibrated tips.
Default priors were used for all parameters except for migration rates between demes for which an exponential prior with mean 1.0 was used.
All analyses were run for 200 million steps across ten independent Markov chains (MCMC runs) and states were sampled every \tbc{XXX} steps.
Due to the complexity of multitype tree parameter space 50\% of states from every analysis were discarded as burn-in, convergence assessed in Tracer v1.6 and states combined using LogCombiner distributed with BEAST v2.4.3 \citep{bouckaert_beast_2014}.
Three chains out of ten did not converge and were discarded altogether.
This left \tbc{XXX} states on which to base posterior inference.

As a secondary analysis to test robustness to choice of prior, we set up an analysis where we increased the mean of the exponential distribution prior for migration rate to 10.0.
All other parameters were identical to the primary analysis and as before 10 independent MCMC chains were run.
In this case, two chains did not converge.
This left \tbc{XXX} states on which to base posterior inference.

As an additional secondary analysis, we also considered a scenario where alignments were split into two fragments (fragment 1 comprised of positions 1-21000, fragment 2 of positions 21000-onwards), with independent clocks, trees and migration rates, but shared substitution models and deme population sizes.
Fragment positions were chosen based on consistent identification of the region around nucleotide 21000 as a probable breakpoint by GARD \citep{pond_gard:_2006} by previous studies into SARS and MERS coronaviruses \citep{hon_evidence_2008,dudas_mers-cov_2016}.
All analyses were set to run for 200 million states, subsampling every 20 000 states.
Chains not converging after 200 million states were discarded.
20\% of the states were discarded as burn-in, convergence assessed with Tracer 1.6 and combined with LogCombiner.
Three chains out of ten did not converge.
This left \tbc{XXX} states on which to base posterior inference.

@@

\subsection*{Demographic inference of MERS-CoV in the reservoir}
In order to infer the demographic history of MERS-CoV in camels we used the results of structured coalescent analyses to identify introductions of the virus into humans.
The oldest sequence from each cluster introduced into humans was used to represent a random draw from the diversity of MERS-CoV circulating in camels.
These sequences were combined with existing sequence data from camels to give us a dataset with minimal demographic signal coming from epidemiological processes in humans.
Sequences belonging to the outgroup clade where most of MERS-CoV sequences from Egypt fall were removed out of concern that MERS epidemics in the Arabian Peninsula and Egypt are distinct epidemics with relatively poor sampling in the latter.
A flexible skygrid tree prior \citep{gill_2013} was used to recover estimates of relative genetic diversity ($N_{e}\tau$) at 50 evenly spaced grid points across six years, ending at the most recent tip in the tree (2015 August) in BEAST v1.8.4 \citep{drummond_bayesian_2012}.
We set up five independent MCMC chains to run for 500 million states, sampling every 50 000 states.
This analysis suffered from poor convergence, where two chains converged onto one stationary distribution, two to another and the last chain onto a third stationary distribution, with high effective sample sizes.
Demographic trajectories recovered by the two main stationary distributions are very similar and differences between the two appear to be caused by convergence onto different tree topologies, with onward effects on evolutionary rate estimates and distributions of coalescence times.
\tbc{This statement seemd a bit strong, as it looked like the $N_{e}\tau$ trajectory was highly similar between these different stationary distributions.}
This non-convergence effect may have been masked previously by the use of all available MERS-CoV sequences from humans which may have lead MCMC towards one of the multiple stationary distributions.

@@

\begin{figure}[h]
\centering
	\includegraphics[width=0.65\textwidth]{figures/mers_skygrid.png}
	\caption{\textbf{Demographic history of MERS-CoV in Arabian peninsula camels.}
Demographic history of MERS-CoV in camels, as inferred via a skygrid coalescent tree prior \citep{gill_2013}.
Two skygrid reconstructions are shown, one for each of the stationary distributions reached by MCMC.
Shaded interval indicates the 95\% highest posterior density interval for the product of generation time and effective population size, $N_{e}\tau$.
Midline tracks the inferred median of $N_{e}\tau$.
Maximum clade credibility (MCC) tree from the skygrid was inferred is shown in the background, with camel sequences highlighted in orange and human sequences highlighted in blue.
	}
	\label{skygrid}
\end{figure}

>>>>

Although we find evidence of seasonality in zoonotic spillover timing, no such relationship exists for sizes of human sequence clusters (figure \ref{seasonality}C? \gdc{note - lettering might change}).
This is entirely expected, since little seasonality in human behaviour that could facilitate MERS-CoV transmission is expected following an introduction.
Similarly, we do not observe any trend in human sequence cluster sizes over time (Figure \ref{seasonality}B, Spearman $\rho = 0.06$, $p=0.68$), suggesting that MERS-CoV outbreaks in humans are neither growing nor shrinking in size.
This is not surprising either, since MERS-CoV is a camel virus that has to date, experienced little-to-no selective pressure to improve transmissibility between humans.

@@

\subsection*{Recombination shapes MERS-CoV diversity}
Recombination has been shown to occur in all genera of coronaviruses, including MERS-CoV \citep{lai_1985,makino_1986,keck_1988,kottier_1995,herrewegh_1998}.
In order to explore the role of recombination in shaping MERS-CoV genetic diversity we used two recombination detection approaches across partitions of taxa corresponding to inferred MERS-CoV clades.
Both methods rely on sampling parental and recombinant alleles within the alignment, although each quantifies different signals of recombination.
One hallmark of recombination is the ability to carry alleles derived via mutation from one lineage to another, which appear as repeated mutations taking place in the recipient lineage, somewhere else in the tree.
The PHI (pairwise homoplasy index) test quantifies the appearance of these excessive repeat mutations (homoplasies) within an alignment \citep{bruen_simple_2006}.
Another hallmark of recombination is spatial clustering of alleles along the genome, due to how template switching, the primary mechanism of recombination in RNA viruses, occurs.
3Seq relies on the spatial structure of nucleotide similarities between sequence triplets -- two potential parent-donors and one candidate offspring-recipient sequences \citep{boni_exact_2007}.

@@

%%% AR - this paragraph is a bit of a mess. Suggests we can't trust most of the analysis above. If we can't believe the camel tree why do we trust a structured coalescent model?
Conversely, our results strongly suggest that co-infection of humans with distinct lineages of MERS-CoV should be exceedingly rare.
We find little evidence that recombination will interfere with the inference of human outbreak clusters (Figure \ref{recombinant_features}A).
Between the two trees in figure \ref{recombinant_features}B four `human' clades are expanded where either singleton introductions or two-taxon clades in fragment 2 join other clades in fragment 1.
For the reverse comparison there are five `human' clades in fragment 2 that are expanded.
All such clades have low divergence (figure \ref{recombinant_features}B) and thus incongruences in human clades are more likely to be caused by differences in deme assignment rather than actual recombination.
And while we observe evidence of distinct phylogenetic trees from different parts of the MERS-CoV genome (Figure \ref{recombinant_features}B), human clades are minimally affected and large portions of the posterior probability density in both parts of the genome are concentrated in shared clades (Figure \ref{flower}).
Critically, we observe the same source-sink dynamics between camel and human populations in trees constructed from separate genomic fragments as were observed in the original full genome tree (Figures \ref{mcc}, \ref{recombinant_features}B).
Observed departures from strictly clonal evolution suggest that while recombination is an issue for inferring MERS-CoV phylogenies, its effect on the human side of MERS outbreaks is minimal, as expected.
MERS-CoV evolution on the reservoir side, though complicated by recombination, is nonetheless still amenable to phylogenetic methods, in part through limited diversity of the virus in camels (see next section).
In humans MERS-CoV evolution should be far easier to track as the only detectable and problematic recombinants are more likely to arise within the transmission chain, than through human co-infection with distinct MERS-CoV lineages.

@@

\begin{figure}[h]
\centering
	\includegraphics[width=0.65\textwidth]{figures/mers_skygrid.png}
	\caption{\textbf{Demographic history of MERS-CoV in Arabian peninsula camels.}
Demographic history of MERS-CoV in camels, as inferred via a skygrid coalescent tree prior \citep{gill_2013}.
Three skygrid reconstructions are shown, red and orange for each of the stationary distributions reached by MCMC with the whole genome and a black one where the genome was split into ten partitions.
Shaded interval indicates the 95\% highest posterior density interval for the product of generation time and effective population size, $N_{e}\tau$.
Midline tracks the inferred median of $N_{e}\tau$.
%Maximum clade credibility (MCC) tree from the skygrid was inferred is shown in the background, with camel sequences highlighted in orange and human sequences highlighted in blue.
	}
	\label{skygrid}
\end{figure}

@@

For our primary analysis, the MultiTypeTree module \citep{vaughan_efficient_2014} of BEAST v2.4.3 \citep{bouckaert_beast_2014} was used to specify a structured coalescent model with two demes -- humans and camels.
This model estimates 4 parameters: rate of coalescence in human viruses, rate of coalescence in camel viruses, rate of migration from the human deme to the camel deme and rate of migration from the camel deme to the human deme.
Analyses were run on codon position partitioned data with two separate HKY+$\Gamma_{4}$ \citep{hky_1985,yang_1994} nucleotide substitution models specified for codon positions 1+2 and 3.
A relaxed molecular clock with branch rates drawn from a lognormal distribution \citep{drummond_2006} was used to infer the evolutionary rate from date calibrated tips.
Default priors were used for all parameters except for migration rates between demes for which an exponential prior with mean 1.0 was used.
All analyses were run for 200 million steps across ten independent Markov chains (MCMC runs) and states were sampled every 20 000 steps.
Due to the complexity of multitype tree parameter space 50\% of states from every analysis were discarded as burn-in, convergence assessed in Tracer v1.6 and states combined using LogCombiner distributed with BEAST v2.4.3 \citep{bouckaert_beast_2014}.
Three chains out of ten did not converge and were discarded altogether.
This left 70 000 states on which to base posterior inference.

As a secondary analysis to test robustness to choice of prior, we set up an analysis where we increased the mean of the exponential distribution prior for migration rate to 10.0.
All other parameters were identical to the primary analysis and as before 10 independent MCMC chains were run.
In this case, two chains did not converge.
This left 80 000 states on which to base posterior inference.

As an additional secondary analysis, we also considered a scenario where alignments were split into two fragments (fragment 1 comprised of positions 1-21000, fragment 2 of positions 21000-29364), with independent clocks, trees and migration rates, but shared substitution models and deme population sizes.
Fragment positions were chosen based on consistent identification of the region around nucleotide 21000 as a probable breakpoint by GARD \citep{pond_gard:_2006} by previous studies into SARS and MERS coronaviruses \citep{hon_evidence_2008,dudas_mers-cov_2016}.
All analyses were set to run for 200 million states, subsampling every 20 000 states.
Chains not converging after 200 million states were discarded.
20\% of the states were discarded as burn-in, convergence assessed with Tracer 1.6 and combined with LogCombiner.
Three chains out of ten did not converge.
This left 70 000 states on which to base posterior inference.

@@

\subsection*{Demographic inference of MERS-CoV in the reservoir}
In order to infer the demographic history of MERS-CoV in camels we used the results of structured coalescent analyses to identify introductions of the virus into humans.
The oldest sequence from each cluster introduced into humans was used to represent a random draw from the diversity of MERS-CoV circulating in camels.
These sequences were combined with existing sequence data from camels to give us a dataset with minimal demographic signal coming from epidemiological processes in humans.
Sequences belonging to the outgroup clade where most of MERS-CoV sequences from Egypt fall were removed out of concern that MERS epidemics in the Arabian Peninsula and Egypt are distinct epidemics with relatively poor sampling in the latter.
A flexible skygrid tree prior \citep{gill_2013} was used to recover estimates of relative genetic diversity ($N_{e}\tau$) at 50 evenly spaced grid points across six years, ending at the most recent tip in the tree (2015 August) in BEAST v1.8.4 \citep{drummond_bayesian_2012}.
We set up five independent MCMC chains to run for 500 million states, sampling every 50 000 states.
This analysis suffered from poor convergence, where two chains converged onto one stationary distribution, two to another and the last chain onto a third stationary distribution, with high effective sample sizes.
Demographic trajectories recovered by the two main stationary distributions are very similar and differences between the two appear to be caused by convergence onto subtly different tree topologies.
This non-convergence effect may have been masked previously by the use of all available MERS-CoV sequences from humans which may have lead MCMC towards one of the multiple stationary distributions.

@@
